---
title: "Vis: Histograms"
author: Zach del Rosario
date: 2020-05-22
output: github_document
time: 30
reading: 20
---

# Vis: Histograms

*Purpose*: *Histograms* are a key tool for EDA. In this exercise we'll get a little more practice constructing and interpreting histograms and densities.

*Reading*: [Histograms](https://rstudio.cloud/learn/primers/3.3)
*Topics*: (All topics)
*Reading Time*: ~20 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

### __q1__ Using the graphs generated in the chunks `q1-vis1` and `q1-vis2` below, answer:

- Which `class` has the most vehicles?
- Which `class` has the broadest distribution of `cty` values?
- Which graph---`vis1` or `vis2`---best helps you answer each question?

```{r q1-vis1}
## NOTE: No need to modify
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_freqpoly(bins = 10)
```


```{r q1-vis2}
## NOTE: No need to modify
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_density()
```

- SUV has the most vehicles (see graph 1).
- Subcompact has the most distribution over cities (see graph 2).

In the previous exercise, we learned how to *facet* a graph. Let's use that part of the grammar of graphics to clean up the graph above.

### __q2__ Modify `q1-vis2` to use a `facet_wrap()` on the `class`. "Free" the vertical axis with the `scales` keyword to allow for a different y scale in each facet.

```{r q2-task}
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_density() +
  facet_wrap(~class, scales = "free_y")
```

In the reading, we learned that the "most important thing" to keep in mind with `geom_histogram()` and `geom_freqpoly()` is to _explore different binwidths_. We'll explore this idea in the next question.

### __q3__ Analyze the following graph; make sure to test different binwidths. What patterns do you see? Which patterns remain as you change the binwidth?

```{r q3-task}
## TODO: Run this chunk; play with differnet bin widths
diamonds %>%
  filter(carat < 1.1) %>%

  ggplot(aes(carat)) +
  geom_histogram(binwidth = 0.05, boundary = 0.005) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.1)

  )
```

**Observations**
- With a bin width of 0.01, it's clear many are exactly 0.5, 0.7, or other round
numbers of carats, but more ``arbitrary" carat numbers are less common.
- With a bin width of 0.1, the most common amount of carats is 0.3 to 0.4 carats.
- When the bin width is significantly less than 0.01, it's clear that most carat
counts are at discrete values and most of the bins are sparse. However the first
observation is still clear.
- Most bin widths can capture the general trend of the data.

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis02-histograms-assignment.Rmd).
